Modelling low Mach number stellar hydrodynamics with MAESTROeX

### 2019-10-04 MAESTROeX_I/appendix.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htb]
\begin{center}
\begin{tabular}{l c}
\multirow{4}{3.25in}[30mm]{ \includegraphics[width=3.0in]{./figs/wdconvect_256_splitU} } & \multicolumn{1}{c}{\includegraphics[width=2.35in]{./figs/magvel_full_XY}} \\
& \multicolumn{1}{c}{\begin{footnotesize} (a) Velocity magnitude, solved using full $\mathbf{U}$ \end{footnotesize}} \\[1.em]
& \multicolumn{1}{c}{\includegraphics[width=2.5in]{./figs/magvel_lineout_X}} \\ 
& \multicolumn{1}{c}{\begin{footnotesize} (b) Comparison along white line \end{footnotesize}} \\
\end{tabular}
\caption{\label{fig:wdconvect_splitU} When solving for the full velocity $\mathbf{U}$ in the nodal projection step instead of the split velocity 
  formulation, we illustrate
  (left) the peak temperature, $T_{\text{peak}}$ in a white dwarf at resolution of $256^3$ during the early evolution of the star and
  (right) the magnitude of velocity at early time. We observe much larger velocities outside the density cutoff region, $\rho < 10^5 \text{ g cm}^{-3}$, 
  that are not seen when we split the velocity dynamics.}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htb]
\begin{center}
\begin{tabular}{l c}
\multirow{4}{3.25in}[30mm]{ \includegraphics[width=3.25in]{./figs/wdconvect_256_splitU} } & \multicolumn{1}{c}{\includegraphics[width=2.75in]{./figs/magvel_full_XY}} \\
& \multicolumn{1}{c}{\begin{footnotesize} (a) Velocity magnitude, solved using full $\mathbf{U}$ \end{footnotesize}} \\[1.em]
& \multicolumn{1}{c}{\includegraphics[width=3.25in]{./figs/magvel_lineout_X}} \\ 
& \multicolumn{1}{c}{\begin{footnotesize} (b) Comparison along white line \end{footnotesize}} \\
\end{tabular}
\caption{\label{fig:wdconvect_splitU} When solving for the full velocity $\mathbf{U}$ in the nodal projection step instead of the split velocity 
  formulation, we illustrate
  (Left) the peak temperature, $T_{\text{peak}}$ in a white dwarf at resolution of $256^3$ during the early evolution of the star and
  (Right) the magnitude of velocity at early time. We observe much larger velocities outside the density cutoff region, $\rho < 10^5 \text{ g cm}^{-3}$, 
  that are not seen when we split the velocity dynamics.}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-10-04 MAESTROeX_II/paper.tex

%% The first piece of markup in an AASTeX v6.x document is the \documentclass
%% command. LaTeX will ignore any data that comes before this command. The 
%% documentclass can take an optional argument to modify the output style.
%% The command below calls the preprint style which will produce a tightly 
%% typeset, one-column, single-spaced document.  It is the default and thus
%% does not need to be explicitly stated.
%%
%%
%% using aastex version 6.3
\documentclass[trackchanges]{aastex63}
%\documentclass{aastex63}

@@

\newcommand{\inp}{\mathrm{in}}
\newcommand{\outp}{\mathrm{out}}
\newcommand{\nph}{{n+\myhalf}}
\newcommand{\nmh}{{n-\myhalf}}
\newcommand{\ow}{\overline{w_0}}
\newcommand{\dw}{\delta w_0}
\newcommand{\uadvone}{\Ub^{\mathrm{ADV},\star}}
\newcommand{\uadvonedag}{\Ub^{\mathrm{ADV},\dagger,\star}}
\newcommand{\uadvtwo}{\Ub^{\mathrm{ADV}}}
\newcommand{\uadvtwodag}{\Ub^{\mathrm{ADV},\dagger}}
\newcommand{\gcc}{\mathrm{g~cm^{-3} }}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% The following section outlines numerous optional output that
%% can be displayed in the front matter or as running meta-data.
%%
%% If you wish, you may supply running head information, although
%% this information may be modified by the editorial offices.
\shorttitle{MAESTROeX Low Mach Number Astrophysics}
\shortauthors{Fan et al.}
%%
%% You can add a light gray and diagonal water-mark to the first page 
%% with this command:
%% \watermark{text}
%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
%% long you can control the water-mark size with:
%% \setwatermarkfontsize{dimension}
%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver}

@@

\begin{abstract}
We present MAESTROeX, a massively parallel solver for low Mach number astrophysical flows.
The underlying low Mach number equation set allows for efficient, long-time integration for highly subsonic flows compared to compressible approaches.
MAESTROeX is suitable for modeling full spherical stars as well as well as planar simulations of dynamics within localized regions of a star,
and can robustly handle several orders of magnitude of density and pressure stratification.
Previously, we have described the development of the predecessor of MAESTROeX, called MAESTRO, in a series of papers. 
Here, we present a new, greatly simplified temporal integration scheme that retains the same order of accuracy as our previous approaches. 
We also explore the use of alternative spatial mapping of the one-dimensional base state onto the full Cartesian grid.
The code leverages the new AMReX software framework for block-structured adaptive mesh refinement (AMR) applications, allowing for scalability to large fractions of leadership-class machines.
Using our previous studies on the convective phase of single-degenerate progenitor models of Type Ia supernovae as a guide, we characterize the performance of the code and validate the new algorithmic features.  Like MAESTRO, MAESTROeX is fully open source.
\end{abstract}

@@

\input{intro.tex}

\input{equations.tex}

\input{algorithm.tex}

\input{results.tex}

\input{conclusions.tex}

\input{appendix.tex}

>>>>

%% The first piece of markup in an AASTeX v6.x document is the \documentclass
%% command. LaTeX will ignore any data that comes before this command. The 
%% documentclass can take an optional argument to modify the output style.
%% The command below calls the preprint style which will produce a tightly 
%% typeset, one-column, single-spaced document.  It is the default and thus
%% does not need to be explicitly stated.
%%
%%
%% using aastex version 6.3
%\documentclass[trackchanges]{aastex63}
\documentclass{aastex63}

@@

\newcommand{\inp}{\mathrm{in}}
\newcommand{\outp}{\mathrm{out}}
\newcommand{\nph}{{n+\myhalf}}
\newcommand{\nmh}{{n-\myhalf}}
\newcommand{\ow}{\overline{w_0}}
\newcommand{\dw}{\delta w_0}
\newcommand{\uadvone}{\Ub^{\mathrm{ADV},\star}}
\newcommand{\uadvonedag}{\Ub^{\mathrm{ADV},\dagger,\star}}
\newcommand{\uadvtwo}{\Ub^{\mathrm{ADV}}}
\newcommand{\uadvtwodag}{\Ub^{\mathrm{ADV},\dagger}}
\newcommand{\gcc}{\mathrm{g~cm^{-3} }}
\newcommand{\kth}{{k_{\mathrm{th}}}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% The following section outlines numerous optional output that
%% can be displayed in the front matter or as running meta-data.
%%
%% If you wish, you may supply running head information, although
%% this information may be modified by the editorial offices.
\shorttitle{SDC Scheme for Low Mach Number Flows}
\shortauthors{Fan et al.}
%%
%% You can add a light gray and diagonal water-mark to the first page 
%% with this command:
%% \watermark{text}
%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
%% long you can control the water-mark size with:
%% \setwatermarkfontsize{dimension}
%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{Spectral Deferred Corrections Scheme for Low Mach Number Flows}

@@

\begin{abstract}
  We describe a spectral deferred corrections scheme for MAESTROeX, an adaptive
  low Mach number hydrodynamics code that allows for efficient, long-time integration. 
\end{abstract}

@@

\section{Introduction} \label{sec:intro}

\section{Numerical Considerations and Algorithm} \label{sec:equations}

MAESTROeX integrates the following system of equations, 
%%
\begin{eqnarray}
\frac{\partial\Ub}{\partial t} &=& -\Ub\cdot\nabla\Ub  - \frac{\beta_0}{\rho}\nabla\left(\frac{\pi}{\beta_0}\right) - \frac{\rho-\rho_0}{\rho} g\eb_r,\label{eq:momentum}\\
\frac{\partial(\rho X_k)}{\partial t} &=& -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:species}\\
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} + \rho\Hnuc + \nabla\cdot\kth\nabla T,\label{eq:enthalpy}
\end{eqnarray}
%%
along with a velocity constraint equation,
%
\begin{equation}
\nabla\cdot(\beta_0\Ub) = \beta_0\left(S - \frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial t}\right), \label{eq:U divergence}
\end{equation}
%
as described in previous papers \citep{MAESTRO_V,MAESTRO_VI}. Note that we have included thermal diffusion
in the governing equations with the term 
\begin{equation}
\nabla\cdot\kth\nabla T = \nabla\cdot\frac{\kth}{c_p}\nabla h - \sum_k\nabla\cdot\frac{\xi_k k_{\rm th}}{c_p}\nabla X_k - \nabla\cdot\frac{h_p k_{\rm th}}{c_p}\nabla p_0 .
\end{equation}
The species are also constrained such that $\sum_k X_k = 1$ giving $\rho = \sum_k (\rho X_k)$ and
%
\begin{equation}
\frac{\partial\rho}{\partial t} = -\nabla\cdot(\rho\Ub).
\end{equation}
%

In past papers, MAESTROeX advance the thermodynamic variables by coupling advection, diffusion,
and reactions using Strang splitting.  Specifically, we integrate the reaction terms 
over half a time step ignoring contributions from advection and diffusion,
then from this intermediate state we integrate the advection and diffusion terms over
a full time step (while ignoring reactions), and finally integrate the reactions over
a half time step (ignoring advection and diffusion). For problems where
the reactions and/or diffusion greatly alter the energy balance or composition 
as compared to advection, this operator splitting approach can lead to large 
splitting errors and highly inaccurate solutions.
This issue is can be particularly exasperating for low Mach number methods that can
theoretically take large advection-based time steps.

An alternate approach to advancing the thermodynamic variables is to use
spectral deferred corrections (SDC). SDC is an iterative scheme used to couple the various processes
together, with each process seeing an  approximation of the other processes as a source term.
The SDC algorithm converges to an integral representation of the solution in
time that couples all of the processes together in a self-consistent fashion.
It has been shown to be more accurate and efficient than Strang splitting in a
terrestrial, non-stratified, low Mach number reacting flow solver \cite{nonaka2012sdc},
so we believe an SDC version of MAESTROeX would greatly improve the efficiency of the code.
\\

\emph{As a first attempt, we will work on coupling advection and reactions only
via SDC, with a base state that is fixed in time, but not space.}
\\

The time-advancement is divided into three major steps.  The first step is the predictor, where we integrate the thermodynamic variables, $(\rho,\rho X_k,\rho h)$, over the full time step.  The second step is corrector, where we use the results from the predictor to perform a more accurate temporal integration of the thermodynamic variables.  The third step is the velocity and dynamic pressure update.\\

{\bf Step 1:} ({\it Compute advection velocities})\\ \\
Use $\Ub^n$ and a second-order Godunov method to compute time-centered edge velocities, $\uadvonedag$, with time-lagged dynamic pressure and explicit buoyancy as forcing terms.  The $\dagger$ superscript indicates that this field does not satisfy the divergence constraint.  Compute $S^{n+\myhalf,\star}$ by extrapolating in time,
\begin{equation}
S^{n+\myhalf,\star} = S^n + \frac{\Delta t^n}{2}\frac{S^n - S^{n-1}}{\Delta t^{n-1}},
\end{equation}
and project $\uadvonedag$ to obtain $\uadvone$, which satisfies
\begin{equation}
\nabla\cdot\left(\beta_0^n\uadvone\right) = S^{n+\myhalf,\star}.
\end{equation}

{\bf Step 2:} ({\it Predictor})\\ \\
 In this step, we integrate $(\rho, \rho X_k, \rho h)$ over the full time step.  The quantities $(S, \beta_0, k_{\rm th}, c_p, \xi_k, h_p)^n$ are computed from the the thermodynamic variables at $t^n$.  This step is divided into several sub-steps:\\

{\bf Step 2A:} ({\it Compute advective flux divergences})\\ \\
Use $\uadvone$ and a second-order Godunov integrator to compute time-centered edge states, $(\rho X_k, \rho h)^{n+\myhalf,(0)}$, with time-lagged reactions ($I^{\rm lagged} = I^{(j_{\rm max})}$ from the previous time step), explicit diffusion, and time-centered thermodynamic pressure as source terms.  Define the advective flux divergences as
\begin{eqnarray}
A_{\rho X_k}^{(0)} &=& -\nabla\cdot\left[\left(\rho X_k\right)^{n+\myhalf,{(0)}}\uadvone\right],\\
A_{\rho h}^{(0)} &=& -\nabla\cdot\left[\left(\rho h\right)^{n+\myhalf,(0)}\uadvone\right] + \uadvone\cdot\nabla p_0.
\end{eqnarray}
Next, use these fluxes to compute the time-advanced density,
\begin{equation}
\frac{\rho^{n+1} - \rho^n}{\Delta t} = \sum_k A_{\rho X_k}^{(0)}.
\end{equation}
Then, compute preliminary, time-advanced species using
\begin{equation}
\frac{\rho^{n+1}\widehat{X}_k^{n+1,(0)} - (\rho X_k)^n}{\Delta t} = A_{\rho X_k}^{(0)} + I_{\rho X_k}^{\rm lagged}.\label{eq:sdc species 2}
\end{equation}

{\bf Step 2B:} ({\it Compute diffusive flux divergence})\\ \\
Solve a Crank-Nicolson-type diffusion equation for $\widehat{h}^{n+1,(0)}$, using transport coefficients evaluated at $t^n$ everywhere,
\begin{eqnarray}
\frac{\rho^{n+1}\widehat{h}^{n+1,(0)} - (\rho h)^n}{\Delta t} &=& A_{\rho h}^{(0)} + I_{\rho h}^{\rm lagged}\nonumber\\
&& + \frac{1}{2}\left(\nabla\cdot\frac{\kth^n}{c_p^n}\nabla h^n + \nabla\cdot\frac{\kth^n}{c_p^n}\nabla \widehat{h}^{n+1,(0)}\right)\nonumber\\
&& - \frac{1}{2}\left(\sum_k\nabla\cdot\frac{\xi_k^n k_{\rm th}^n}{c_p^n}\nabla X_k^n + \sum_k\nabla\cdot\frac{\xi_k^n k_{\rm th}^n}{c_p^n}\nabla\widehat{X}_k^{n+1,(0)}\right)\nonumber\\
&& - \frac{1}{2}\left(\nabla\cdot\frac{h_p^n k_{\rm th}^n}{c_p^n}\nabla p_0 + \nabla\cdot\frac{h_p^n k_{\rm th}^n}{c_p^n}\nabla p_0\right),\label{eq:sdc enthalpy 2}
\end{eqnarray}

{\bf Step 2C:} ({\it Advance thermodynamic variables})\\ \\
Define $Q_{\rho X_k}^{(0)}$ as the right hand side of (\ref{eq:sdc species 2}) without the $I_{\rho X_k}^{\rm lagged}$ term, and define $Q_{\rho h}^{(0)}$ as the right hand side of (\ref{eq:sdc enthalpy 2}) without the $I_{\rho h}^{\rm lagged}$ term.  Use {\tt VODE} to integrate (\ref{eq:species}) and (\ref{eq:enthalpy}) over $\Delta t$ to advance $(\rho X_k, \rho h)^n$ to $(\rho X_k, \rho h)^{n+1,(0)}$ using the piecewise-constant advection and diffusion source terms:
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& Q_{\rho X_k}^{(0)} + \rho\dot\omega_k\\
\frac{\partial(\rho h)}{\partial t} &=& Q_{\rho h}^{(0)} + \rho\Hnuc.
\end{eqnarray}
At this point we can define $I_{\rho X_k}^{(0)}$ and $I_{\rho h}^{(0)}$, or whatever term we need depending on our species and enthalpy edge state prediction types, for use in the corrector step.  In our first implementation, we are predicting $\rho X_k$ and $\rho h$, in which case we define:
\begin{eqnarray}
I_{\rho X_k}^{(0)} &=& \frac{(\rho X_k)^{n+1,(0)} - (\rho X_k)^n}{\Delta t} - Q_{\rho X_k}^{(0)}\\
I_{\rho h}^{(0)} &=& \frac{(\rho h)^{n+1,(0)} - (\rho h)^n}{\Delta t} - Q_{\rho h}^{(0)}.
\end{eqnarray}

{\bf Step 3:} ({\it Update advection velocities})\\ \\
First, compute $S^{n+\myhalf}$ and $\beta_0^{n+\myhalf}$ using
\begin{equation}
S^{n+\myhalf} = \frac{S^n + S^{n+1,(0)}}{2}, \qquad \beta_0^{n+\myhalf} = \frac{\beta_0^n + \beta_0^{n+1,(0)}}{2}.
\end{equation}
Then, project $\uadvtwodag$ to obtain $\uadvtwo$, which satisfies
\begin{equation}
\nabla\cdot\left(\beta_0^{n+\myhalf}\uadvtwo\right) = S^{n+\myhalf}.
\end{equation}

{\bf Step 4:} ({\it Corrector Loop})\\ \\
We loop over this step from $j=1,j_{\rm max}$ times.  In the corrector, we use the time-advanced state from the predictor to perform a more accurate integration of the thermodynamic variables.  The quantities $(S, \beta_0, k_{\rm th}, c_p, \xi_k, h_p)^{n+1,(j-1)}$ are computed from $(\rho,\rho X_k,\rho h)^{n+1,(j-1)}$.  This step is divided into several sub-steps:\\

{\bf Step 4A:} ({\it Compute advective flux divergences})\\ \\
Use $\uadvtwo$ and a second-order Godunov integrator to compute time-centered edge states, $(\rho X_k, \rho h)^{n+\myhalf}$, with iteratively-lagged reactions ($I^{(j-1)}$), explicit diffusion, and time-centered thermodynamic pressure as source terms.  Define the advective flux divergences as
\begin{eqnarray}
A_{\rho X_k}^{(j)} &=& -\nabla\cdot\left[\left(\rho X_k\right)^{n+\myhalf,(j)}\uadvtwo\right],\\
A_{\rho h}^{(j)} &=& -\nabla\cdot\left[\left(\rho h\right)^{n+\myhalf,(j)}\uadvtwo\right] + \uadvtwo\cdot\nabla p_0.
\end{eqnarray}
Then, compute preliminary, time-advanced species using
\begin{equation}
\frac{\rho^{n+1}\widehat{X}_k^{n+1,(j)} - (\rho X_k)^n}{\Delta t} = A_{\rho X_k}^{(j)} + I_{\rho X_k}^{(j-1)}.\label{eq:sdc species 3}
\end{equation}

{\bf Step 4B:} ({\it Compute diffusive flux divergence})\\ \\
Solve a backward-Euler-type correction equation for $\widehat{h}^{n+1,(j)}$,
\begin{eqnarray}
\frac{\rho^{n+1}\widehat{h}^{n+1,(j)} - (\rho h)^n}{\Delta t} &=& A_{\rho h}^{(j)} + I_{\rho h}^{(j-1)}\nonumber\\
&& + \nabla\cdot\frac{\kth^{n+1,(j-1)}}{c_p^{n+1,(j-1)}}\nabla\widehat{h}^{n+1,(j)} + \frac{1}{2}\left(\nabla\cdot\frac{\kth^n}{c_p^n}\nabla h^n - \nabla\cdot\frac{\kth^{n+1,(j-1)}}{c_p^{n+1,(j-1)}}\nabla h^{n+1,(j-1)}\right)\nonumber\\
&& - \frac{1}{2}\left(\sum_k\nabla\cdot\frac{\xi_k^n k_{\rm th}^n}{c_p^n}\nabla X_k^n + \sum_k\nabla\cdot\frac{\xi_k^{n+1,(j-1)} k_{\rm th}^{n+1,(j-1)}}{c_p^{n+1,(j-1)}}\nabla\widehat{X}_k^{n+1,(j)}\right)\nonumber\\
&& - \frac{1}{2}\left(\nabla\cdot\frac{h_p^n k_{\rm th}^n}{c_p^n}\nabla p_0 + \nabla\cdot\frac{h_p^{n+1,(j-1)}k_{\rm th}^{n+1,(j-1)}}{c_p^{n+1,(j-1)}}\nabla p_0\right),\label{eq:sdc enthalpy 3}
\end{eqnarray}

{\bf Step 4C:} ({\it Advance thermodynamic variables})\\ \\
Define $Q_{\rho X_k}^{(j)}$ as the right hand side of (\ref{eq:sdc species 3}) without the $I_{\rho X_k}^{(j-1)}$ term, and define $Q_{\rho h}^{(j)}$ as the right hand side of (\ref{eq:sdc enthalpy 3}) without the $I_{\rho h}^{(j-1)}$ term.  Use {\tt VODE} to integrate (\ref{eq:species}) and (\ref{eq:enthalpy}) over $\Delta t$ to advance $(\rho X_k, \rho h)^n$ to $(\rho X_k, \rho h)^{n+1,(j)}$ using the piecewise-constant advection and diffusion source terms:
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& Q_{\rho X_k}^{(j)} + \rho\dot\omega_k\\
\frac{\partial(\rho h)}{\partial t} &=& Q_{\rho h}^{(j)} + \rho\Hnuc.
\end{eqnarray}
At this point we can define $I_{\rho X_k}^{(j)}$, $I_{\rho h}^{(j)}$, and any other $I$ terms we need depending on 
our species and enthalpy edge state prediction types, for use in the predictor in the next time step.  In our first implementation, we are predicting $\rho X_k$ and $\rho h$, in which case we define:
\begin{eqnarray}
I_{\rho X_k}^{(j)} &=& \frac{(\rho X_k)^{n+1,(j)} - (\rho X_k)^n}{\Delta t} - Q_{\rho X_k}^{(j)}\\
I_{\rho h}^{(j)} &=& \frac{(\rho h)^{n+1,(j)} - (\rho h)^n}{\Delta t} - Q_{\rho h}^{(j)}.
\end{eqnarray}

{\bf Step 5:} ({\it Advance velocity and dynamic pressure})\\ \\
Similar to the original MAESTROeX algorithm, we first compute the face-centered, time-centered velocities, $\Ub^{\nph,\pred}$
using a Godunov approach \citep{XRB_III}. Then, we update
the velocity field $\Ub^n$ to $\Ub^{n+1,\dagger}$ by discretizing
equation (\ref{eq:momentum}) as
\begin{equation}
\Ub^{n+1,\dagger}
= \Ub^n - \dt \left[\uadvtwo \cdot \nabla \Ub^{\nph,\pred} \right]
 - \dt \left[ \frac{\beta_0^\nph}{\rho^\nph} \nabla \left( \frac{\pi^\nmh}{\beta_0^\nmh} \right) + \frac{\left(\rho^\nph-\rho_0^\nph\right)}{\rho^\nph} g^{\nph} \eb_r \right],
\end{equation}
where
\begin{equation}
\rho^\nph = \frac{\rho^n + \rho^{n+1}}{2}, \qquad \rho_0^\nph = \frac{\rho_0^n + \rho_0^{n+1}}{2}.
\end{equation}
Again, the $\dagger$ superscript refers
to the fact that the updated velocity does not satisfy the divergence constraint,
\begin{equation}
\nabla \cdot \left(\beta_0^{n+1} \Ub^{n+1} \right) = \beta_0^{n+1} \left[ S^{n+1} - \frac{1}{\gammaonebar^{n+1}p_0^{n+1}}\left(\frac{\partial p_0}{\partial t}\right)^{\nph}\right].\label{eq:div3}
\end{equation}
We use an approximate projection to project $\Ub^{n+1,\dagger}$ onto the space of velocities that satisfy the constraint to obtain $\Ub^{n+1}$ using a ``nodal'' projection.
This projection necessarily differs from the MAC projection used in
{\bf Step 1} and {\bf Step 3} because the velocities in those steps are defined
on faces and $\Ub^{n+1}$ is defined at cell centers, requiring different divergence
and gradient operators..

\section{Numerical Algorithm} \label{sec:algorithm}

\section{Results} \label{sec:results}

### 2019-10-04 astronum_2019_MAESTROeX/astronum-maestroex.tex

\begin{abstract}
Modelling long-time convective flows in the interiors of stars is extremely challenging using conventional compressible hydrodynamics codes due to the acoustic timestep limitation.
Many of these flows are in the low Mach number regime, which allows us to exploit the relationship between acoustic and advective time scales to develop a more computationally efficient approach.
\maestroex\ is an open source low Mach number stellar hydrodynamics code that allows much larger timesteps to be taken, therefore enabling systems to be modelled for much longer periods of time. This is particularly important for the problem of convection in the cores of rotating massive stars prior to core collapse. To fully capture the dynamics, it is necessary to model these systems in three dimensions at high resolution over many rotational periods.  We present an overview of \maestroex's current capabilities, describe ongoing work to incorporate the effects of rotation and discuss how we are optimizing the code to run on GPUs. 
\end{abstract}

@@

Sound-proof methods for modelling low Mach number flows can take a variety of forms. One technique is to use preconditioners in order to reduce the stiffness of the equations, allowing larger timesteps to be used \cite{Miczek2014,Barsukow2016}. Another technique is to modify the flow parameters, artificially boosting the speed of the flow without changing its behavior so that the system evolves faster and fewer timesteps are required \citep{Rempel2005,Hotta2012}. Fully implicit time integration codes, such as those used by \cite{Viallet2011,Viallet2015,Goffrey2016}, are no longer restricted by the CFL condition and so can use arbitrarily large timesteps.

@@

We have begun porting \maestroex~to GPUs, leveraging the GPU capabilities built into \amrex~and following the lessons learned porting \castro~to GPUs. In the \amrex~framework, routines can be ported to GPUs by simply inserting a few macros into the code. The boilerplate code required to compile this for GPU is then generated by a custom preprocessor prior to compilation. CUDA managed memory is used to take care of data transfers between the CPU and GPU, so no explicit copies of data to/from the CPU and GPU are required, significantly reducing the extra code required to port the code. One of the key emphases of the \amrex~approach to GPUs is to maintain performance portability: porting code to GPUs should not come at the expense of the code's performance on CPU-only machines. Because GPU-specific code is created by the preprocessor prior to compilation, it is not necessary to maintain separate versions of the code for different architectures. By changing the compiler flags, the same bit of code can be compiled for serial, MPI-only, OpenMP-only, MPI+OpenMP, GPU-only or MPI+GPU.
In order to achieve the best performance on the GPU, we require some additional optimization of the computational kernels. However, we find that these optimizations often improve the performance when the code is run on the CPU only as well. 
More details of the approach used to port our codes to GPU are to appear in a forthcoming \amrexastro~GPU paper.

@@

As mentioned, our work porting \maestroex~to run on GPUs is still ongoing, however in Figure~\ref{fig:gpu_speedup} we show the speedup we have achieved so far for a number of individual functions that have been offloaded to GPU. The plot shows the average execution time of each function, recorded for the three-dimensional \texttt{reacting\_bubble} problem on the OLCF Summit machine.
This problem uses a simple reaction network modelling $^{12}\mathrm{C} + {}^{12}\mathrm{C} \rightarrow {}^{24}\mathrm{Mg}$ integrated on the GPUs using the CUDA Fortran port of VODE \citep{vode} described in \cite{Zingale2018}.
Both tests were run on a single node, consisting of two IBM POWER9 processors (together providing 42 physical CPU cores) and six NVIDIA Volta GPUs. The CPU test used 42 MPI processes, each with 4 OpenMP threads, and the GPU test used 6 MPI processes, each process associated with a single GPU.
Note that the plot's $x$-axis scale is logarithmic. As can be seen, large speedups in excess of 30-40$\times$ were achieved for several of the functions when run on the GPU. Once we have finished porting the remaining routines to GPU and further optimised our code, we hope to achieve this sort of speedup for the entire timestep.

>>>>

\begin{abstract}
Modelling long-time convective flows in the interiors of stars is extremely challenging using conventional compressible hydrodynamics codes due to the acoustic timestep limitation.
Many of these flows are in the low Mach number regime, which allows us to exploit the relationship between acoustic and advective time scales to develop a more computationally efficient approach.
\maestroex\ is an open source low Mach number stellar hydrodynamics code that allows much larger timesteps to be taken, therefore enabling systems to be modelled for much longer periods of time. This is particularly important for the problem of convection in the cores of rotating massive stars prior to core collapse. To fully capture the dynamics, it is necessary to model these systems in three dimensions at high resolution over many rotational periods.  We present an overview of \maestroex's current capabilities, describe ongoing work to incorporate the effects of rotation and discuss how we are optimising the code to run on GPUs. 
\end{abstract}

@@

Sound-proof methods for modelling low Mach number flows can take a variety of forms. One technique is to use preconditioners in order to reduce the stiffness of the equations, allowing larger timesteps to be used \cite{Miczek2014,Barsukow2016}. Another technique is to modify the flow parameters, artificially boosting the speed of the flow without changing its behaviour so that the system evolves faster and fewer timesteps are required \citep{Rempel2005,Hotta2012}. Fully implicit time integration codes, such as those used by \cite{Viallet2011,Viallet2015,Goffrey2016}, are no longer restricted by the CFL condition and so can use arbitrarily large timesteps.

@@

We have begun porting \maestroex~to GPUs, leveraging the GPU capabilities built into \amrex~and following the lessons learned porting \castro~to GPUs. In the \amrex~framework, routines can be ported to GPUs by simply inserting a few macros into the code. The boilerplate code required to compile this for GPU is then generated by a custom preprocessor prior to compilation. CUDA managed memory is used to take care of data transfers between the CPU and GPU, so no explicit copies of data to/from the CPU and GPU are required, significantly reducing the extra code required to port the code. One of the key emphases of the \amrex~approach to GPUs is to maintain performance portability: porting code to GPUs should not come at the expense of the code's performance on CPU-only machines. Because GPU-specific code is created by the preprocessor prior to compilation, it is not necessary to maintain separate versions of the code for different architectures. By changing the compiler flags, the same bit of code can be compiled for serial, MPI-only, OpenMP-only, MPI+OpenMP, GPU-only or MPI+GPU.
In order to achieve the best performance on the GPU, we require some additional optimisation of the computational kernels. However, we find that these optimisations often improve the performance when the code is run on the CPU only as well. 
More details of the approach used to port our codes to GPU are to appear in a forthcoming \amrexastro~GPU paper.

@@

As mentioned, our work porting \maestroex~to run on GPUs is still ongoing, however in Figure~\ref{fig:gpu_speedup} we show the speedup we have achieved so far for a number of individual functions that have been offloaded to GPU. The plot shows the average execution time of each function, recorded for the three-dimensional \texttt{reacting\_bubble} problem on the OLCF Summit machine.
This problem uses a simple reaction network modelling $^{12}\mathrm{C} + {}^{12}\mathrm{C} \rightarrow {}^{24}\mathrm{Mg}$ that can be integrated on the GPUs using the CUDA Fortran port of VODE \citep{vode} described in \cite{Zingale2018}.
Both tests were run on a single node, consisting of two IBM POWER9 processors (together providing 42 physical CPU cores) and six NVIDIA Volta GPUs. The CPU test used 42 MPI processes, each with 4 OpenMP threads, and the GPU test used 6 MPI processes, each process associated with a single GPU.
Note that the plot's $x$-axis scale is logarithmic. As can be seen, large speedups in excess of 30-40$\times$ were achieved for several of the functions when run on the GPU. Once we have finished porting the remaining routines to GPU and further optimised our code, we hope to achieve this sort of speedup for the entire timestep.
